Bioconductor workflow for single-cell RNA sequencing: Normalization, dimensionality reduction, clustering, and lineage inference

### 2017-12-04 workflow/workflow.Rmd
update Rcode and EP comments

@@

@@

@@

@@

@@

EP: We could add back in the mergeClusters step now that we don't have the W in the data slot. I haven't done that yet because we want to just recreate the previous, but it would be reasonable now. It would be based on doing DE of the normalized values, however, because I put them into the main assay slot, so we might need to think about this multiple assay question more seriously.  But it would be nice to show that you can do that for this setting. Once this is running I can come back to it.

@@

```{r heatmapsClusters, fig.cap="RSEC: Heatmap of the normalized expression measures for the 1,000 most variable genes, where rows correspond to genes and columns to cells ordered by RSEC clusters."}
# Set colors for cell clusterings
experimentColors <- bigPalette[1:nlevels(colData(dataObj)$Experiment)]
batchColors <- bigPalette[1:nlevels(colData(dataObj)$Batch)]
metaColors <- list("Experiment" = experimentColors,
                   "Batch" = batchColors)

@@

```{r boxplotReg3g, fig.width=6, fig.height=4, fig.cap="RSEC: Boxplots of the log count of gene Reg3g stratified by cluster."}
# c4 <- rep("other clusters", ncol(se))
# c4[primaryClusterNamed(dataObj) == "c4"] <- "cluster c4"
# boxplot(log1p(assay(se)["Reg3g", ]) ~ primaryClusterNamed(dataObj),
#         ylab = "Reg3g log counts", cex.axis = .8, cex.lab = .8)
plotFeatureBoxplot(dataObj, whichCluster="primary", feature="Reg3g")
```

@@

```{r fitgam}
t <- pseudotime(lineages)[,1]
y <- assay(pseudoCe)
gam.pval <- apply(y,1,function(z){
  d <- data.frame(z=z, t=t)
  tmp <- gam(z ~ lo(t), data=d)
  p <- summary(tmp)[4][[1]][1,5]
  p
})
```

EP: Since changed NormalizedValues to the main assay, this cleans up the code above and below.

```{r heatmapsignificant, fig.cap="DE: Heatmap of the normalized expression measures for the 100 most significantly DE genes for the neuronal lineage, where rows correspond to genes and columns to cells ordered by pseudotime."}
topgenes <- names(sort(gam.pval, decreasing = FALSE))[1:100]
orderSamples(pseudoCe)<-order(t,na.last=TRUE)
plotHeatmap(pseudoCe[,!is.na(t)], clusterSamplesData = "orderSamplesValue", breaks = .99)
```

>>>>

EP: I have only updated the code, not the text. The text needs adaptation, especially future directions and references to `SummarizedExperiment` which should be `SingleCellExperiment`.

@@

EP: Should update these.

@@

EP: Note that clusterExperiment uses `parallel` package, so might should update this.

@@

EP: There's code in SingleCellExperiment for spike-ins. We should use it, though I'm not familiar with it.

@@

EP: Should probably address these normalized values, especially to more strongly make clear that shouldn't use these for clustering, just for visualization. The way its stated it sounds like dimensionality reduction is an "also", when in fact it is the main point and the normalized values are the afterthought. Might consider reordering the sections to more clearly demonstrate this and put dimensionality reduction first.

@@

EP: We could add back in the mergeClusters step now that we don't have the W in the data slot. I haven't done that yet because we want to just recreate the previous, but it would be reasonable now. It would be based on doing DE of the normalized values, however, because I put them into the main assay slot, so we might need to think about this multiple assay question more seriously. But then again, I don't know the workflow for the DE part, so maybe we want to leave it out.  But it would be nice to show that you can do that for this setting. Once this workflow is running I can come back to it.

@@

```{r heatmapsClusters, fig.cap="RSEC: Heatmap of the normalized expression measures for the 1,000 most variable genes, where rows correspond to genes and columns to cells ordered by RSEC clusters."}
# Set colors for additional sample data
experimentColors <- bigPalette[1:nlevels(colData(dataObj)$Experiment)]
batchColors <- bigPalette[1:nlevels(colData(dataObj)$Batch)]
metaColors <- list("Experiment" = experimentColors,
                   "Batch" = batchColors)

@@

```{r boxplotReg3g, fig.width=6, fig.height=4, fig.cap="RSEC: Boxplots of the log count of gene Reg3g stratified by cluster."}
# c4 <- rep("other clusters", ncol(se))
# c4[primaryClusterNamed(dataObj) == "c4"] <- "cluster c4"
# boxplot(log1p(assay(se)["Reg3g", ]) ~ primaryClusterNamed(dataObj),
#         ylab = "Reg3g log counts", cex.axis = .8, cex.lab = .8)
plotFeatureBoxplot(dataObj, whichCluster="primary", feature="Reg3g",ylab = "Reg3g log counts", cex.axis = .8, cex.lab = .8)
```

@@

```{r fitgam}
t <- pseudotime(lineages)[,1]
y <- transformData(pseudoCe)
gam.pval <- apply(y,1,function(z){
  d <- data.frame(z=z, t=t)
  tmp <- gam(z ~ lo(t), data=d)
  p <- summary(tmp)[4][[1]][1,5]
  p
})
```

EP: Since changed NormalizedValues to the main assay, this cleans up the code above and below. I've used the `transformData` command, which will not do anything if isCount=FALSE above, but will convert to right if we change isCount=TRUE. It's better to use that command than `assay` for that reason. Might should add comment in code or text explaining this.

```{r heatmapsignificant, fig.cap="DE: Heatmap of the normalized expression measures for the 100 most significantly DE genes for the neuronal lineage, where rows correspond to genes and columns to cells ordered by pseudotime."}
topgenes <- names(sort(gam.pval, decreasing = FALSE))[1:100]
orderSamples(pseudoCe)<-order(t,na.last=TRUE)
plotHeatmap(pseudoCe[topgenes,!is.na(t)], clusterSamplesData = "orderSamplesValue", breaks = .99)
```

EP: note the above with do heirarchcial clustering of the top genes. If want them to stay in order given, should give `clusterFeaturesData=topgenes` (I think that is the argument)
